Supplemental Information: Aligning marine species range data to better serve science and conservation

### 2016-09-11 case_study_mpa/3_mpa_spp_analysis.Rmd
all text updated, including references.  Still need to complete abstract.

mpa_all_runs1 <- mpa_all_runs %>%
  group_by(trial) %>%
  mutate(n_trial = n()) %>%
  group_by(trial, cat, n_trial) %>%
  summarize(p_cat = n()/first(n_trial)) %>%
  ungroup() %>%
  mutate(cat = factor(cat, levels = c('gap', '0-2%', '2-5%', '5-10%', '>10%'))) %>%
  mutate(trial = factor(trial, levels = c('IUCN 2015', 'AM 2015 (0%) +\nIUCN 2015', 'AM 2015 (0%)', 'AM 2015 (50%)', 'AM 2014 (50%)'))) %>%
  mutate(p_cat = p_cat * 100) %>%
  arrange(cat)

@@

### Plot the bar chart
barchart_mpa_gap <- ggplot(mpa_all_runs1, 
                           aes(x = trial, fill = cat, weight = p_cat)) +
  ggtheme_plot + 
  theme(panel.grid.major.y = element_blank()) +
  geom_bar(stat = 'count', alpha = 1) +
  scale_fill_manual(values = c('>10%'  = '#4dac26', 
                               '5-10%' = '#b8e186', 
                               '2-5%'  = '#fbe5c5', 
                               '0-2%'  = '#f1b6da',
                               'gap'   = '#d01c8b')) +
  ### add grid lines; horizontal but then get flipped
  geom_hline(yintercept = break_nums, size = 0.25, color = 'white', alpha = .5) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 125), breaks = break_nums,
                     labels = sprintf('%s%%', break_nums)) + 
  geom_text(aes(label = sprintf('n = %s', n_trial), y = 101), hjust = 0, 
            size = 2, 
            color = 'grey30') +
  scale_x_discrete(labels = c('5. IUCN 2015',
                              '4. AM 2015 (0%) +\nIUCN 2015', 
                              '3. AM 2015 (0%)', 
                              '2. AM 2015 (50%)', 
                              '1. AM 2014 (50%)')) +
  coord_flip() +
  labs(x = 'Scenario', 
       y = 'Percent of species by MPA coverage', 
       fill = 'MPA coverage')

>>>>

mpa_all_runs1 <- mpa_all_runs %>%
  group_by(trial) %>%
  mutate(n_trial = n()) %>%
  group_by(trial, cat, n_trial) %>%
  summarize(p_cat = n()/first(n_trial)) %>%
  ungroup() %>%
  mutate(cat = factor(cat, levels = c('gap', '0-2%', '2-5%', '5-10%', '>10%'))) %>%
  filter(trial != 'AM 2015 (0%) +\nIUCN 2015') %>%
  mutate(trial = factor(trial, levels = c('IUCN 2015', 
                                          # 'AM 2015 (0%) +\nIUCN 2015', 
                                          'AM 2015 (0%)', 'AM 2015 (50%)', 'AM 2014 (50%)'))) %>%
  mutate(p_cat = p_cat * 100) %>%
  arrange(cat)

@@

### Plot the bar chart
barchart_mpa_gap <- ggplot(mpa_all_runs1, 
                           aes(x = trial, fill = cat, weight = p_cat)) +
  ggtheme_plot + 
  theme(panel.grid.major.y = element_blank()) +
  geom_bar(stat = 'count', alpha = 1) +
  scale_fill_manual(values = c('>10%'  = '#4dac26', 
                               '5-10%' = '#b8e186', 
                               '2-5%'  = '#fbe5c5', 
                               '0-2%'  = '#f1b6da',
                               'gap'   = '#d01c8b')) +
  ### add grid lines; horizontal but then get flipped
  geom_hline(yintercept = break_nums, size = 0.25, color = 'white', alpha = .5) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 125), breaks = break_nums,
                     labels = sprintf('%s%%', break_nums)) + 
  geom_text(aes(label = sprintf('n = %s', n_trial), y = 101), hjust = 0, 
            size = 2, 
            color = 'grey30') +
  scale_x_discrete(labels = c('4. IUCN 2015',
                              # '4. AM 2015 (0%) +\nIUCN 2015', 
                              '3. AM 2015 (0%)', 
                              '2. AM 2015 (50%)', 
                              '1. AM 2014 (50%)')) +
  coord_flip() +
  labs(x = 'Scenario', 
       y = 'Percent of species by MPA coverage', 
       fill = 'MPA coverage')

### 2016-09-11 ms_1_intro.Rmd
all text updated, including references.  Still need to complete abstract.

The two most comprehensive, widely-used global-scale repositories that predict marine species ranges throughout the world's oceans are AquaMaps, which rely primarily on model predictions [@kaschner_aquamaps:_2015], and range data from the International Union for Conservation of Nature (IUCN), which rely primarily on expert opinion [@iucn_iucn_2015].  The two datasets are designed to serve different purposes:  IUCN range maps are drawn to estimate extent of occurrence (EOO) [@GastonXXX 1994], communicating range size as an [indicator/factor/criterion] for extinction risk assessments [@IUCNXXX]; AquaMaps essentially communicate the area of occupancy (AOO) [@REFXXX] of a species based on habitat suitability [@AquaMapsXXX].

The fundamental differences between the datasets (EOO versus AOO, measure of range size versus distribution) suggest that the choice of one over the other should be carefully matched to the purpose for which it is to be used [@GastonXXX 2008, XXX Other refs], and yet these datasets have been used in hundreds of studies and applications for a wide range of purposes, including assessing marine species status [@halpern_index_2012; @halpern_patterns_2015; @selig_assessing_2013-1], evaluating global biodiversity patterns [@coll_biodiversity_2010; @martin_manual_2014; @pimm_biodiversity_2014; @kaschner_current_2011], predicting species range shifts [@garciamolinos_climate_2015], and setting conservation priorities [@klein_shortfalls_2015].  In most of these cases, the implications of choosing one versus the other of these datasets is not evaluated or discussed, yet strong conservation and management conclusions are drawn from the results of the studies.

<!--
_The fundamental differences between the intent of the two datasets precludes casual interchangeability between the two [@GastonXXX 2008], and yet they_

_"The two measures (EOO and AOO) serve different purposes, and should not be regarded as alternatives that simply differ in accuracy" - Gaston et al 2004 - literature suggests that this distinction precludes casual interchangeability between the two [@gastonXXX 2008]; they have specific purposes.  And yet they get used, with no explicit acknowldegment of the difference, for a wide range of conservation studies.  It's important to understand the implications and measures to help adapt them to specific uses._
-->

Here we focus not on how the two datasets should ideally be used; instead, in recognizing that each provides value for many aspects of conservation research, we highlight hidden assumptions and sources of error within each dataset, and explore the implications of choosing one over the other.

Uncertainties inherent in method and intent inevitably drive differences in the range predictions made by each dataset: extent of occurrence range data such as IUCN range maps will inevitably introduce commission errors (inaccurate indication of presence), while species distribution models such as AquaMaps are expected to introduce fewer commission errors at the expense of more omission errors (inaccurate indications of absence) [@rondinini_tradeoffs_2006].  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range [@rondinini_tradeoffs_2006; @jetz_ecological_2008].

To understand the implications of differences between the AquaMaps and IUCN datasets, we compare how each data source represents the global spatial and taxonomic distribution of the `r format(n_spp_all, big.mark = ',')` marine species mapped by one or both datasets.  Most notably, AquaMaps includes range maps for many more species (currently `r format(n_spp_am, big.mark = ',')` species; `r p_spp_am`% of total), such that most global analyses related to marine biodiversity to date have used AquaMaps (IUCN range map data exist for only `r format(n_spp_iucn, big.mark = ',')` unique marine species).  Yet the small overlap in mapped species between the two datasets means each provides unique value in taxonomic and geographic coverage.

Although relatively small in number, these overlap species present a unique opportunity to evaluate the two datasets overall.  As understanding of a given species improves, we expect that range maps, regardless of method, would more closely approach the true distribution in predicting species presence and absence.  Close alignment between two independent range maps can therefore indicate higher confidence in species presence or absence than a single map.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align in both spatial distribution and overall area.

Using this alignment framework, we explore methodological issues that can introduce errors, and suggest methods to improve confidence in species range predictions.  In particular, we show how explicitly using depth as a constraint on coral ranges in IUCN range data dramatically mitigates a source of commission error, and highlight a side effect of AquaMaps methodology that results in potentially unexpected species range boundaries.

@@

As a case study, we then reexamine a global analysis of gaps in protection afforded by marine protected areas (MPAs) [@klein_shortfalls_2015] as a case study to explore the implications of selecting one data set over the other.  Repeating the study using IUCN range data rather than AquaMaps data increases the apparent proportion of gap species from 1.4% to 6.4%, while also increasing the proportion of well-protected species from 1.5% to 2.9%.  This shift highlights the implications of different data use decisions on our understanding of marine biodiversity status and protection.

While we have identified some potential issues in each dataset, we cannot simply recommend one data set over the other.  Instead, we recommend simple methods to account for incidence of errors in each dataset, improving confidence in species range predictions to better inform conservation management and policy decisions.

>>>>

The two most comprehensive, widely-used global-scale repositories that predict marine species ranges throughout the world's oceans are AquaMaps, which rely primarily on model predictions to communicate the distribution of a species based on habitat suitability [@kaschner_aquamaps:_2015], and range data from the International Union for Conservation of Nature (IUCN), which rely primarily on expert opinion to communicate range size as a criterion for extinction risk assessments [@iucn_iucn_2015].

While the two datasets ostensibly describe the same information, i.e. where can a particular species be found, they communicate fundamentally distinct concepts.  The IUCN range map for a given species describes the "extent of occurrence" for that species, a contiguous region that encloses all known occurrences and connecting regions, with the explicit caveat that this "does not mean that it is distributed equally within that polygon or occurs everywhere within that polygon" [@iucn_iucn_2015].  The AquaMaps map for a species, on the other hand, seems to better be defined as "area of occupancy": the subset of the extent of occurrence in which the species is actually likely to be found, for example by excluding unsuitable habitats from the extent of occurrence [@iucn_iucn_2015].  The distinctions between extent of occurrence and area of occupancy, including methodology, limitations, and appropriate usage, are well documented [@rondinini_tradeoffs_2006; @gaston_sizes_2009]. ___include reference to a supplemental table for more details?___

<!--
_paragraph describing IUCN and AM as extent of occurrence and area of occupancy respectively, with basic defs and implication that AM will generally be smaller than IUCN; possibly refer to Table S1 for defs and details?_ _one reference: http://www.biodiversitya-z.org/content/species-extent-of-occurrence-eoo "The intent behind this parameter is to measure the degree to which risks from threatening factors are spread spatially across the taxon’s geographical distribution. It is not intended to be an estimate of the amount of occupied or potential habitat, or a general measure of the taxon’s range" and http://www.biodiversitya-z.org/content/species-area-of-occupancy-aoo "The measure reflects the fact that a taxon will not usually occur throughout the area of its extent of occurrence, which may, for example, contain unsuitable habitats"_
-->

The fundamental differences between the datasets suggest that the choice of one over the other should be carefully matched to the purpose for which it is to be used [@gaston_sizes_2009], and yet these datasets have been used in hundreds of studies and applications for a wide range of purposes, including assessing marine species status [@halpern_index_2012; @halpern_patterns_2015; @selig_assessing_2013-1], evaluating global biodiversity patterns [@coll_biodiversity_2010; @martin_manual_2014; @pimm_biodiversity_2014; @kaschner_current_2011], predicting species range shifts [@garciamolinos_climate_2015], and setting conservation priorities [@klein_shortfalls_2015].  In most of these cases, the implications of choosing one versus the other of these datasets is not evaluated or discussed, yet strong conservation and management conclusions are drawn from the results of the studies.

Here we focus not on how the two datasets should ideally be used; instead, recognizing that each provides distinct value for conservation research, we explore hidden assumptions and sources of error within each dataset, and test the implications of choosing one over the other.  You do science with the data you have, not with the data you wish you had; often, the ideal type of spatial data is not available, and non-ideal data is better than no data at all.  In such a case, it is crucial to acknowledge, and where possible address, the resulting tradeoffs.

___include as table, in body or in supplemental?  It really kind of doesn't fit here any more___ _Uncertainties inherent in method and intent inevitably drive differences in the range predictions made by each dataset: extent of occurrence range data such as IUCN range maps will inevitably introduce commission errors (inaccurate indication of presence), while species distribution models such as AquaMaps are expected to introduce fewer commission errors at the expense of more omission errors (inaccurate indications of absence) [@rondinini_tradeoffs_2006].  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range [@rondinini_tradeoffs_2006; @jetz_ecological_2008]._

Data availability is a crucial factor in deciding which dataset to use.  We compare how each data source represents the global spatial and taxonomic distribution of the `r format(n_spp_all, big.mark = ',')` marine species mapped by one or both datasets.  Most notably, AquaMaps includes range maps for many more species (currently `r format(n_spp_am, big.mark = ',')` species; `r p_spp_am`% of total), such that most global analyses related to marine biodiversity to date have used AquaMaps (IUCN range map data exist for only `r format(n_spp_iucn, big.mark = ',')` unique marine species).  Yet the small overlap in mapped species between the two datasets means each provides unique value in taxonomic and geographic coverage.

Although relatively small in number, these overlap species present a unique opportunity to evaluate the two datasets overall.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align in both spatial distribution and overall area.  Based upon the definitions of extent of occurrence and area of occupancy, we expect that for a given species, the AquaMaps predicted distribution will fall within, and describe a smaller range than, the IUCN predicted distribution.  Where these expectations seem to fail, we explore methodological issues that can introduce errors.  In particular, we show how explicitly using depth as a constraint on coral ranges in IUCN range data dramatically mitigates a source of commission error, and highlight a side effect of AquaMaps methodology that results in discontinuous species range boundaries.

<!--As understanding of a given species improves, we expect that range maps, regardless of method, would more closely approach the true distribution in predicting species presence and absence.  Close alignment between two independent range maps can therefore indicate higher confidence in species presence or absence than a single map.-->

@@

As a case study, we then reexamine a global analysis of gaps in protection afforded by marine protected areas (MPAs) [@klein_shortfalls_2015] as a case study to explore the implications of selecting one data set over the other.  Repeating the study using IUCN range data rather than AquaMaps data quintuples the apparent proportion of "gap" species, i.e. those species whose ranges remain completely unprotected, from 1.2% to 6.4%.  This shift highlights the implications of different data use decisions on our understanding of marine biodiversity status and protection.

Due to the variation in georegional and taxonomic coverage, and fundamental distinctions in method and intent of these two spatial range datasets, there is no way to simply recommend one dataset over the other.  <!--Instead, we recommend simple methods to account for incidence of errors in each dataset, accounting for and reconciling differences in species range predictions to better inform conservation management and policy decisions.-->  Instead, by making explicit the distinctions between the two, and offering methods to identify and reconcile potential errors in each, we hope to reinforce the effective use of these datasets to better inform conservation management and policy decisions.

### 2016-09-11 ms_2_methods.Rmd
all text updated, including references.  Still need to complete abstract.

The IUCN publishes species range maps developed by species experts.  These experts outline spatial boundaries that define the "limits of distribution" of a given species, based on observation records and informed by expert understanding of species' range and habitat preferences (Fig S1A).  <!--The result is essentially an extent of occurrence (EOO) map for the species-->

In contrast, AquaMaps models species distribution based on envelopes of environmental preference, such as temperature, depth, and salinity; these preference envelopes are deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a global 0.5° grid to determine suitable habitat, resulting in a "probability of occurrence" for each species (Fig S1B).  To roughly constrain species ranges to appropriate georegions, the AquaMaps model uses Food and Agriculture Organization of the United Nations (FAO) Major Fishing Area [@REF] boundaries.  Of the resulting maps, 1296 (5.7%) have been further refined through an expert review process [@kaschner_mapping_2006; @ready_predicting_2010].

<!--In addition to these methodological differences, the two datasets also differ in intent.  IUCN maps essentially communicate the extent of occurrence (EOO) while AquaMaps uses suitable habitat as an indicator of probability of occurrence, -->

@@

__Comparison of taxonomic and regional distribution__:
To examine the overall taxonomic distribution across the spatial datasets, we grouped species by taxonomic class and data source, and determined the proportion of each class represented in each dataset.  To compare the spatial representation of the two datasets directly, we rasterized the IUCN species polygons to the same 0.5° grid as the AquaMaps species maps; we determined species presence within a grid cell as any non-zero overlap of a species polygon with the cell  ___(Fig S2)___.  For the AquaMaps dataset, we determined per-cell species count by including all species with non-zero probability of occurrence, to best approximate the "extent of occurrence" generally indicated by IUCN maps.

__Alignment framework for paired maps__: OR __Comparison of paired maps__:
Species range maps communicate a prediction of both the distribution of a species and the size of a species range.  To enable direct comparison of these measures between the two datasets, we identified "paired map" species, i.e. the subset of marine species that have range maps in both IUCN and AquaMaps current native distribution, using genus and species binomials as a matching key.  We used the taxize package [@REF] in R [@REF] to standardize species names and synonyms; for species with separate subpopulation maps in IUCN, we combined all subpopulations to create a single global population.   For each of these paired map species, we determined species presence within each spatial cell for each dataset using the same criteria outlined above.

@@

__(some heading here? to separate from alignment)__
Corals provide a useful case study to examine differences in species range alignment.  They are an intensely studied taxonomic group whose importance in supporting biodiversity is undisputed; their dependence on photosynthesis provides habitat-limiting life history traits, and they don't tend to move.  Extracting data from the IUCN API [@REF?] we identified depth limitations of each coral species mapped in the IUCN dataset to verify that none are found below the photosynthetic depth limit of 200 m.  Using the same methodology as shown in Fig 2, we created a 200 m bathymetry raster from Natural Earth's 200 m bathymetry polygon [@natural_earth_XXX] and masked our IUCN coral rasters to identify mapped coral presence below 200 m.  The resulting maps were again compared to the AquaMaps rasters to examine distribution alignment and area ratio.

@@

The oceans abound with rare and poorly-understood species who offer sparse information on behavior and habitat preferences to inform models and expert opinion.  To identify such data-poor species within the set of paired map species, we used the ___robis???___ and ___rgbif???___ packages [@REFS] in R to identify known occurrences of each species; occurrences were averaged between the two occurrence databases as a proxy for data richness.

@@

## Methods for MPA Gap Analysis case study
To assess the effectiveness of MPAs in protecting biodiversity, Klein et al. [-@klein_shortfalls_2015] compared the coverage of the global MPA network presented by the World Database on Protected Areas (WDPA) [@iucn_world_2014] to the species ranges described in the 2014 AquaMaps dataset [@aquamaps_XXX].  For the primary analysis, the researchers defined species presence as 50% or greater probability of occurrence.

>>>>

The IUCN publishes species range maps developed by species experts.  These experts outline spatial boundaries that define the "limits of distribution" of a given species, based on observation records and informed by expert understanding of species' range and habitat preferences (Fig S1A).  ___some note about IUCN directions for creating the maps???___.  For the purposes of this analysis, we did not use IUCN data for bird species, which are available separately through Bird Life International [@birdlife_international_and_natureserve_bird_2015].

In contrast, AquaMaps models species distribution based on envelopes of environmental preference, such as temperature, depth, and salinity; these preference envelopes are deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a global 0.5° grid to determine suitable habitat, resulting in a "probability of occurrence" for each species (Fig S1B).  To roughly constrain species ranges to appropriate georegions, the AquaMaps model uses Food and Agriculture Organization of the United Nations (FAO) Major Fishing Area [@fao_fisheries_and_aquaculture_department_cwp_2015] boundaries.  Of the resulting maps, 1296 (5.7%) have been further refined through an expert review process [@kaschner_mapping_2006; @ready_predicting_2010].

@@

__Comparison of taxonomic and regional distribution:__
To examine the overall taxonomic distribution across the spatial datasets, we grouped species by taxonomic class and data source, and determined the proportion of each class represented in each dataset.  To compare the spatial representation of the two datasets directly, we rasterized the IUCN species polygons to the same 0.5° grid as the AquaMaps species maps; we determined species presence within a grid cell as any non-zero overlap of a species polygon with the cell  ___(Fig S2)___.  For the AquaMaps dataset, we determined per-cell species count by including all species with non-zero probability of occurrence, to best approximate the "extent of occurrence" generally indicated by IUCN maps.

__Comparison of paired maps:__
Species range maps communicate a prediction of both the distribution of a species and the size of a species range.  To enable direct comparison of these measures between the two datasets, we identified "paired map" species, i.e. the subset of marine species that have range maps in both IUCN and AquaMaps current native distribution, using genus and species binomials as a matching key.  We used the taxize package [@scott_chamberlain_taxize_2013] in R [@r_core_team_r:_2016] to standardize species names and synonyms; for species with separate subpopulation maps in IUCN, we combined all subpopulations to create a single global population.   For each of these paired map species, we determined species presence within each spatial cell for each dataset using the same criteria outlined above.

@@

__Examining issues in paired map alignment:__
Corals provide a useful case study to examine differences in species range alignment.  Corals are an intensely studied taxonomic group whose importance in supporting biodiversity is undisputed; their dependence on photosynthesis provides habitat-limiting life history traits, and they don't tend to move.  Extracting data from the IUCN API [@iucn_iucn_2015] we identified depth limitations of each coral species mapped in the IUCN dataset to verify that none are found below the photosynthetic depth limit of 200 m.  Using the same methodology as shown in Fig 2, we created a 200 m bathymetry raster from a 200 m bathymetry spatial dataset (public domain; available from Natural Earth, www.naturalearthdata.com) and masked our IUCN coral rasters to identify mapped coral presence below 200 m.  The resulting maps were again compared to the AquaMaps rasters to examine distribution alignment and area ratio.

@@

The oceans abound with rare and poorly-understood species which offer sparse information on behavior and habitat preferences to inform models and expert opinion.  To identify such data-poor species within the set of paired map species, we used R clients for OBIS [@obis_data_2016] and GBIF [@gbif_global_2010] to identify known occurrences of each species; occurrences were averaged between the two occurrence databases as a proxy for data richness.

@@

## Methods for MPA Gap Analysis case study
To assess the effectiveness of MPAs in protecting biodiversity, Klein et al. [-@klein_shortfalls_2015] compared the coverage of the global MPA network presented by the World Database on Protected Areas (WDPA) [@iucn_world_2014] to the species ranges described in the 2014 AquaMaps dataset [@kaschner_aquamaps:_2015].  For the primary analysis, the researchers defined species presence as 50% or greater probability of occurrence.
